---
title: "Online Text Exploration"
author: "Bastiaan Quast"
date: "Monday, November 10, 2014"
output:
  md_document
bibliography: bibliography.bib
---

# Abstract
We analyse three corpora of US English text found online.
We find that the **blogs** and **news** corpora are similar,
the **twitter** corpus is different.
We propose that this is the result of the 140 character limit of Twitter messages.

# Introduction
In this report we look at three corpora of US English text, a set of internet blogs posts, a set of internet news articles, and a set of twitter messages.

We collect the following forms of information:

 1. file size
 2. number of lines
 3. number of non-empty lines
 4. number of words
 5. distribution of words (quantiles and plot)
 6. number of characters
 7. number of non-white characters
 
In the following section we will describe the data collection process,
the section after that gives the results of the data exploration,
we finally present conclusions and give references.

For our analysis we use the R computing environment [@R], as well as the libraries **stringi** [@stringi] and **ggplot2** [@ggplot2].
In order to make the code more readable we use the pipe operator from the **magrittr** library [@magrittr].
This report is compiled using the **rmarkdown ** library [@rmarkdown] and [@knitr].
Finally during writing we used the RStudio IDE [@RStudio].

# Data

The data is presented as a [ZIP compressed archive](http://en.wikipedia.org/wiki/Zip_(file_format)), which is freely downloadable from [here](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip).
```{r}
# specify the source and destination of the download
destination_file <- "Coursera-SwiftKey.zip"
source_file <- "http://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip"

# execute the download
download.file(source_file, destination_file)

# extract the files from the zip file
unzip(destination_file)
```

Inspect the unzipped files
```{r}
# find out which files where unzipped
unzip(destination_file, list = TRUE )

# inspect the data
list.files("final")
list.files("final/en_US")
```

The corpora are contained in three separate plain-text files,
out of which one is binary, for more information on this see [@newtest].
We import these files as follows.
```{r}
# import the blogs and twitter datasets in text mode
blogs <- readLines("final/en_US/en_US.blogs.txt", encoding="UTF-8")
twitter <- readLines("final/en_US/en_US.twitter.txt", encoding="UTF-8")

# import the news dataset in binary mode
con <- file("final/en_US/en_US.news.txt", open="rb")
news <- readLines(con, encoding="UTF-8")
close(con)
rm(con)
```

Full instructions for importing the data can be found in the [CodeBook](https://github.com/bquast/Data-Science-Capstone/blob/master/CodeBook.md) of the [GitHub repository](https://github.com/bquast/Data-Science-Capstone).


# Basic Statistics

The before we analyse the files we look at their size (presented in MegaBytes / MBs).
```{r}
# file size (in MegaBytes/MB)
file.info("final/en_US/en_US.blogs.txt")$size   / 1024^2
file.info("final/en_US/en_US.news.txt")$size    / 1024^2
file.info("final/en_US/en_US.twitter.txt")$size / 1024^2
```

For our analysis we need two libraries.
```{r}
# library for character string analysis
library(stringi)

# library for plotting
library(ggplot2)
```

We analyse the lines and characters.
```{r}
stri_stats_general( blogs )
stri_stats_general( news )
stri_stats_general( twitter )
```

Next we count the words per item (line). We summarise the distibution of these counts per corpus, using summary statistics and a distibution plot. we start with the **blogs** corpus.
```{r}
words_blogs   <- stri_count_words(blogs)
summary( words_blogs )
qplot(   words_blogs )
```

Next we analys the **news** corpus.
```{r}
words_news    <- stri_count_words(news)
summary( words_news )
qplot(   words_news )
```

Finally we analyse the **twitter** corpus.
```{r}
words_twitter <- stri_count_words(twitter)
summary( words_twitter )
qplot(   words_twitter )
```

# Conclusions
We analyse three corpora of US english text. The file sizes are around 200 MegaBytes (MBs) per file.

We find that the **blogs** and **news** corpora consist of about 1 million items each,
and the *twitter** corpus consist of over 2 million items.
Twitter messages have a character limit of 140 (with exceptions for links),
this explains why there are some many more items for a corpus of about the same size.

This result is further supported by the fact that the number of characters is similar for all three corpora (around 200 million each).

Finally we find that the frequency distributions of the **blogs** and **news ** corpora are similar (appearing to be log-normal).
The frequency distribution of the **twitter** corpus is again different, as a result of the character limit.


# References
